---
title: "DALEX"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: "Learn to explore, explain and examine classification models with DALEX using the Titanic dataset."
---

```{r setup, include=FALSE, message=FALSE}
library(learnr)
library(archivist)
library(DALEX)
library(rms)
library(randomForest)
library(gbm)
library(e1071)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)

# Data
titanic  <- archivist::aread("pbiecek/models/27e5c")
johnny_d <- archivist::aread("pbiecek/models/e3596")
henry    <- archivist::aread("pbiecek/models/a6538")
 
# Train models
titanic_lrm <- archivist::aread("pbiecek/models/58b24")
titanic_rf  <- archivist::aread("pbiecek/models/4e0fc")
titanic_gbm <- archivist::aread("pbiecek/models/b7078")
titanic_svm <- archivist::aread("pbiecek/models/9c27f")

# Create explainer objects
titanic_lrm_exp <- explain(titanic_lrm,
                           data = titanic[, -9],
                           y = titanic$survived == "yes",
                           label = "Logistic Regression",
                           type = "classification",
                           verbose = FALSE)
titanic_rf_exp <- explain(model = titanic_rf, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Random Forest",
                       verbose = FALSE)
titanic_gbm_exp <- explain(model = titanic_gbm, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Generalized Boosted Regression",
                       verbose = FALSE)
titanic_svm_exp <- explain(model = titanic_svm, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Support Vector Machine",
                       verbose = FALSE)
# 
# # Create model performance object
# mp_lms <- model_performance(exp_lms)
# mp_rf <- model_performance(exp_rf)
# 
# # Instances
# joe <- titanic_imputed[1,]
# lucy <- data.frame(
#       class = factor("1st", levels = c("1st", "2nd", "3rd",
#                      "deck crew", "engineering crew",
#                      "restaurant staff", "victualling crew")),
#       gender = factor("female", levels = c("female", "male")),
#       age = 18, sibsp = 0, parch = 0, fare = 70,
#       embarked = factor("Southampton", levels = c("Belfast",
#                         "Cherbourg","Queenstown","Southampton")))
# 
# bd_joe <- predict_parts(exp_rf, joe, type = "break_down_interactions")
# bd_lucy <- predict_parts(exp_rf, lucy, type = "break_down_interactions")
# sh_joe_rf <- predict_parts(exp_rf, joe, type = "shap")
# sh_joe_lms <- predict_parts(exp_lms, joe, type = "shap")
```

## 1 Welcome

This tutorial will teach you how to explore, explain and examine predictive models using the [DALEX](https://modeloriented.github.io/DALEX/) package in R. DALEX is short for mo<b>D</b>el <b>A</b>gnostic <b>L</b>anguage for <b>E</b>xploration and e<b>X</b>planation.

### 1.1 Learning objectives

We will cover:    

+ How to obtain insight into the overall behavior of a model.
+ How to obtain insight into a model's prediction for a single observation.
+ How to examine model performance and understanding the weaknesses.

We will only use _model-agnostic_ methods. These methods treat machine learning models as black boxes. They do not need access to model internals. They work by changing the input of the model and measuring changes in the prediction output.

This tutorial focuses solely on explainability of classification models. However, DALEX can just as easily be used to explain regression models.

This tutorial is based on the book [Explanatory Model Analysis](https://pbiecek.github.io/ema/) by Przemyslaw Biecek and Tomasz Burzykowski. It is a great resource and highly recommended. You can read it for free online. The book will also be available in print from early 2021.

### 1.2 Pre-requisites 
We assume you have basic knowledge of R, specifically of training machine learning models.

Let's get started!

## 2 Titanic dataset
We will use the `titanic` dataset. It contains the details of a subset of persons aboard the Titanic when it sank: 

* *gender*, the person's gender;
* *age*, the person's age in years; 
* *class*, the class in which the passenger travelled, or the duty class of a crew member;
* *embarked*, the harbor in which the person embarked on the ship;
* *country*, person’s home country;
* *fare*, the price of the ticket;
* *sibsp*, the number of siblings/spouses aboard the ship;
* *parch*, the number of parents/children aboard the ship;
* *survived*, the person's survival of the disaster, _yes_ or _no_.

Click **Run Code** to see the first six rows of the dataset.   

```{r ex1, exercise = TRUE}
head(titanic)
```

In the next steps of this tutorial we will create two different kind of models. They both predict the probability of survival aboard the Titanic. So `survived` is the dependent variable. 

## 3 Models

### 3.1 Logistic regression model

It makes sense to start with a logistic regression model. We use restricted cubic splines to model the non-linear effect of age on the odds of survival.     

Click **Run Code** to train the model.

```{r ex2, exercise = TRUE}
library(rms)
titanic_lrm <- lrm(survived == "yes" ~ class + gender + rcs(age) + sibsp + parch + fare + embarked, 
                   data = titanic)
cat("The logistic regression model has been created.")
```

### 3.2 Random forest model

We also create a random forest model. Insert the name of the dependent variable in the code chunk. Run the code. Click the **Solution** button to get help.

```{r ex3, exercise = TRUE}
library(randomForest)
set.seed(1313)
titanic_rf <- randomForest(___ ~ class + gender + age + sibsp + parch + fare + embarked, 
                           data = titanic)
cat("The random forest model has been created.")
```

```{r ex3-solution}
library(randomForest)
set.seed(1313)
titanic_rf <- randomForest(survived ~ class + gender + age + sibsp + parch + fare + embarked, 
                           data = titanic)
cat("The random forest model has been created.")
```

### 3.3 Gradient boosting model

Additionally, we create a gradient boosting model. This may take some time.

```{r ex4, exercise = TRUE, exercise.timelimit = 120}
library(gbm)
set.seed(1313)
titanic_gbm <- gbm(survived == "yes" ~ class + gender + age + sibsp + parch + fare + embarked, 
                   data = titanic, 
                   n.trees = 15000, 
                   distribution = "bernoulli")
cat("The gradient boosting model has been created.")
```

### 3.4 Support vector machine model

Finally, we also create a support vector machine (SVM) model. 

```{r ex5, exercise = TRUE}
library(e1071)
set.seed(1313)
titanic_svm <- svm(survived == "yes" ~ class + gender + age + 
                 sibsp + parch + fare + embarked, data = titanic, 
                 type = "C-classification", probability = TRUE)
cat("The support vector machine model was created.")
```


## 4 Models' predictions

Let's predict the probability of survival for two passengers aboard the Titanic: Johnny D and Henry. 

### 4.1 Johnny D

For convenience, a dataset with Johnny D's data has already been created.

```{r ex6, exercise = TRUE, exercise.eval = TRUE}
print(johnny_d)
```

As we can see, Johnny D is an 8-year-old boy who embarked in Southampton and travelled in the first class with no parents nor siblings, and with a ticket costing 72 pounds.

We obtain the predicted probability of survival for Johnny D for the different models with the generic function `predict()`.

```{r ex7, exercise = TRUE}
pred_lrm <- predict(titanic_lrm, johnny_d, type = "fitted")
pred_rf  <- predict(titanic_rf,  johnny_d, type = "prob")
pred_gbm <- predict(titanic_gbm, johnny_d, type = "response", n.trees = 15000)
pred_svm <- predict(titanic_svm, johnny_d, probability = TRUE)
cat("Predicted probability of survival for Johnny D\n",
    "==============================================\n",
    sprintf("Logistic regression model: %.*f\n", 2, pred_lrm),
    sprintf("Random forest model: %.*f\n", 2, pred_rf[1, "yes"]),    
    sprintf("Gradient boosting model: %.*f\n", 2, pred_gbm), 
    sprintf("Support vector machine model: %.*f\n", 2, attr(pred_svm, "probabilities")[1,"TRUE"]), 
    sep = "")
```

The predictions differ substantially. What causes these differences? Which model deserves our trust? Let's try and find the answers to these questions!

### 4.2 Henry

Henry is an 47-year-old man who embarked in Cherbourg and travelled in the first class with no parents nor siblings, and with a ticket costing 25 pounds.

```{r ex8, exercise = TRUE, exercise.eval = TRUE}
print(henry)
```

For Henry we also predict the odds of survival using different models.

```{r ex9, exercise = TRUE}
pred_lrm <- predict(titanic_lrm, henry, type = "fitted")
pred_rf  <- predict(titanic_rf,  henry, type = "prob")
pred_gbm <- predict(titanic_gbm, henry, type = "response", n.trees = 15000)
pred_svm <- predict(titanic_svm, henry, probability = TRUE)
cat("Predicted probability of survival for Henry\n",
    "==============================================\n",
    sprintf("Logistic regression model: %.*f\n", 2, pred_lrm),
    sprintf("Random forest model: %.*f\n", 2, pred_rf[1, "yes"]),    
    sprintf("Gradient boosting model: %.*f\n", 2, pred_gbm), 
    sprintf("Support vector machine model: %.*f\n", 2, attr(pred_svm, "probabilities")[1,"TRUE"]), 
    sep = "")
```

```{r q1}
question("How does the prediction for Henry compare to that for Johnny D?",
  answer("Henry's predicted probability of survival is <i>lower</i>.", correct = TRUE),
  answer("The predicted probabilities of survival for Henry and Johnny D are <i>equal</i>."),
  answer("Henry's predicted probability of survival is <i>higher</i>.")
)
```

## 5 Models' explainers

The models trained in the previous step have different internal structures and interfaces. We need a unified interface, so that we can compare the models. For this purpose we create _model explainer objects_. An explainer object contains all elements needed to create a model explanation.

### 5.1 Create an explainer object

We use the function `explain()` from the `DALEX` package to create an explainer object. The most frequently used function arguments are:

* `model` (required), the model to be explained;
* `data`, data to which the model is to be applied, be sure to remove the dependent variable; 
* `y`, observed values of the dependent variable corresponding to the data given in the `data` object;
* `label`, unique name of the model;
* `type`, information about the type of the model, either `classification` or `regression`,
* `verbose`, logical indicating whether diagnostic messages are to be printed.

Let's start with an explainer for the logistic regression model. Complete the code block below and run it.

```{r ex10, exercise = TRUE}
library(DALEX)
titanic_lrm_exp <- explain(___,
                           data = titanic[, -9],
                           y = ___$survived == "yes",
                           label = "Logistic Regression",
                           type = ___)
```

```{r ex10-solution}
library(DALEX)
titanic_lrm_exp <- explain(titanic_lrm,
                           data = titanic[, -9],
                           y = titanic$survived == "yes",
                           label = "Logistic Regression",
                           type = "classification")
```

Once the explainer has been created, we can retrieve basic information about the model.

```{r ex11, exercise = TRUE}
titanic_lrm_exp$model_info
```

We also create explainer objects for the other models.

``````{r ex12, exercise = TRUE}
titanic_rf_exp <- explain(model = titanic_rf, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Random Forest",
                       verbose = FALSE)
titanic_gbm_exp <- explain(model = titanic_gbm, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Generalized Boosted Regression",
                       verbose = FALSE)
titanic_svm_exp <- explain(model = titanic_svm, 
                       data = titanic[, -9],
                       y = titanic$survived == "yes", 
                       label = "Support Vector Machine",
                       verbose = FALSE)
cat("The explainers have been created.")
```

Get the information for the support vector machine model from the appropriate explainer.

```{r ex13, exercise = TRUE}

```

```{r ex13-solution}
titanic_svm_exp$model_info
```

```{r q2}
l <- unlist(strsplit(titanic_svm_exp$model_info$ver, "[.]"))
create_answer <- function(){
  index <- sample(1:3, 1)
  l[index] <<- as.character(as.integer(l[index]) + 1)
  return(paste(l, collapse = "."))
}
question("What version of the <code>e1071</code> package was used to create the support vector machine model?",
         answer(create_answer()),
         answer(titanic_svm_exp$model_info$ver, correct = TRUE),  
         answer(create_answer()),
         answer(create_answer())
)
```
